---
title: "Growth in the OA subset of PubMed Central 2000-2013"
author: "Neil Saunders"
date: "28/11/2014"
output:
  github_document:
    toc: true
  html_document:
    toc: yes
---

# Introduction

This document is a second response to [Contrary to what you read in Nature, Open Access has not caused the growth in science publishing](http://www.michaeleisen.org/blog/?p=1654).

Brief summary: I agree with the conclusions, but still not with the numbers :)

# Data
This time, we'll stick with PubMed Central.

We search PMC between the years 2000-2013 as follows:

* Total number of articles published in a year using the PDAT (date of publication) qualifier _e.g._ "2000[PDAT]"
* Filter for the open access (OA) subset using the filter "open+access[FILT]"

```{r message=FALSE}
library(rentrez)
library(ggplot2)
library(reshape2)
```

```{r cache=TRUE}
# create list to hold data
d <- data.frame(year = NA, all = NA, free = NA)
# count total and free articles for 2000-2013
for(i in 1:14) {
    e.all  <- entrez_search("pmc", paste(i+1999, "[PDAT]", sep = ""))
    e.free <- entrez_search("pmc", paste(i+1999, "[PDAT] AND open+access[FILT]", sep = ""))
    d[i, "year"] <- i+1999
    d[i, "all"]  <- e.all$count
    d[i, "free"] <- e.free$count
}
```

Now we calculate the proportion of OA articles and the cumulative sum (year on year increase) of both all and OA articles.

```{r cache=TRUE}
d$prop    <- d$free / d$all
d$all.cs  <- cumsum(d$all)
d$free.cs <- cumsum(d$free)
d
```

# Analysis

##Free full text articles as proportion of all articles

The OA subset articles rose steadily from around 11% of all articles (2000) to 59% (2013).

```{r tidy=TRUE}
ggplot(d) + geom_bar(aes(year, all), stat = "identity", fill = "blue") + geom_bar(aes(year, free), stat = "identity", fill = "orange") + theme_bw() + ylab("number of articles") + xlab("year") + scale_x_continuous(breaks = seq(2000, 2013, by = 2)) + labs(title = "Number of PMC articles per year 2000-2013\nblue = all; orange = OA subset")
```

## Cumulative increase in free full text articles as proportion of cumulative increase in all articles

42 171 PMC articles were published in 2000. The cumulative sum of all articles published between then and 2013 is 1 881 571.

The yearly increase of OA articles is somewhat less than that of all articles.

```{r tidy=TRUE}
ggplot(d) + geom_bar(aes(year, all.cs), stat = "identity", fill = "blue") + geom_bar(aes(year, free.cs), stat = "identity", fill = "orange") + theme_bw() + ylab("cumulative sum of articles") + xlab("year") + scale_x_continuous(breaks = seq(2000, 2013, by = 2)) + labs(title = "Cumulative sum of PMC articles 2000-2013\nblue = all; orange = OA subset")
```

## Cumulative increase in articles relative to start year

The rate of accumulation of OA articles compared with the base year of 2000 is much higher than that for all articles.

```{r tidy=TRUE, cache=TRUE}
d2 <- data.frame(year = 2000:2013, all = d$all.cs / d$all.cs[1], free = d$free.cs / d$free.cs[1])
d2.m <- melt(d2, id.vars = "year")
ggplot(d2.m) + geom_point(aes(year, value, color = variable)) + theme_bw() + scale_color_manual(values = c("blue", "orange")) + scale_x_continuous(breaks = seq(2000, 2013, by = 2)) + ylab("cumulative sum of articles relative to 2000") + labs(title = "Cumulative sum of PMC articles 2000-2013 relative to 2000\nblue = all; orange = OA subset")
```

## Change in number of articles as a proportion of previous year

The year-on-year change in number of articles is not very informative, as there is a lot of unexplained fluctuation.

```{r cache=TRUE, tidy=TRUE}
d3 <- data.frame(year = NA, all = NA, free = NA)
for(i in 1:13) {
    d3[i, "year"] <- d[i+1, "year"]
    d3[i, "all"]  <- (d[i+1, "all"] - d[i, "all"]) / d[i, "all"]
    d3[i, "free"] <- (d[i+1, "free"] - d[i, "free"]) / d[i, "free"]
}
d3.m <- melt(d3, id.vars = "year")
ggplot(d3.m) + geom_line(aes(year, value, color = variable)) + theme_bw() + scale_color_manual(values = c("blue", "orange")) + scale_x_continuous(breaks = seq(2000, 2013, by = 2)) + ylab("relative change in number of articles") + labs(title = "Change in number of PMC articles relative to previous year\nblue = all; orange = OA subset")
```

# Conclusions

From 2000-2013:

* Total and open access articles in PMC rose rapidly
* The proportion of OA articles rose from around 11% to 59%
* Total articles accumulated at a slightly higher rate than OA articles
* From 2005 onwards, OA articles accumulated at a much higher rate than total articles relative to a baseline year of 2000
* Year-on-year changes in article numbers show no clear pattern
* ~~Definition of "open access" for this purpose is not easy :)~~
