---
title: "Growth in free and closed scientific publications 2000-2013"
author: "Neil Saunders"
date: "28/11/2014"
output:
  github_document:
    toc: true
  html_document:
    toc: yes
---

# Introduction

This document is in response to [Contrary to what you read in Nature, Open Access has not caused the growth in science publishing](http://www.michaeleisen.org/blog/?p=1654).

**Note:** I made an error in this analysis by assuming that the PubMed "freetext" filter was a good approximation for open-access. This is not the case; I should have used the OA subset of PubMed Central and did so in a second report, also available here.

# Data

We search PubMed between the years 2000-2013 as follows:

* Total number of articles published in a year using the PDAT (date of publication) qualifier _e.g._ "2000[PDAT]"
* Filter by "free full text available"" using the filter "freetext[FILT]"

Note that other filters are available: "free not pmc" and "free only pmc". However, "freetext" would seem to be the most appropriate to capture all "freely-available" articles.

```{r message=FALSE}
library(rentrez)
library(ggplot2)
library(reshape2)
```

```{r cache=TRUE, tidy=TRUE, }
e.test <- data.frame(year = 2000, free.only.pmc = entrez_search("pubmed", "2000[PDAT] AND free+only+pmc[FILT]")$count, free.not.pmc = entrez_search("pubmed", "2000[PDAT] AND free+not+pmc[FILT]")$count, freetext = entrez_search("pubmed", "2000[PDAT] AND freetext[FILT]")$count)
e.test
```

```{r cache=TRUE}
# create list to hold data
d <- data.frame(year = NA, all = NA, free = NA)
# count total and free articles for 2000-2013
for(i in 1:14) {
    e.all  <- entrez_search("pubmed", paste(i+1999, "[PDAT]", sep = ""))
    e.free <- entrez_search("pubmed", paste(i+1999, "[PDAT] AND freetext[FILT]", sep = ""))
    d[i, "year"] <- i+1999
    d[i, "all"]  <- e.all$count
    d[i, "free"] <- e.free$count
}
```

Now we calculate the proportion of free full text articles and the cumulative sum (year on year increase) of both all and free full text articles.

```{r cache=TRUE}
d$prop    <- d$free / d$all
d$all.cs  <- cumsum(d$all)
d$free.cs <- cumsum(d$free)
d
```

# Analysis

## Free full text articles as proportion of all articles

The "freetext" articles rose steadily from around 21% of all articles (2000) to 33% (2013).

```{r tidy=TRUE}
ggplot(d) + geom_bar(aes(year, all), stat = "identity", fill = "blue") + geom_bar(aes(year, free), stat = "identity", fill = "orange") + theme_bw() + ylab("number of articles") + xlab("year") + scale_x_continuous(breaks = seq(2000, 2013, by = 2)) + labs(title = "Number of PubMed articles per year 2000-2013\nblue = all; orange = freetext")
```

## Cumulative increase in free full text articles as proportion of cumulative increase in all articles

529 860 PubMed articles were published in 2000. The cumulative sum of all articles published between then and 2013 is 10 934 738.

It is clear from this chart that the yearly increase of freetext articles is much less than that of all articles.

```{r tidy=TRUE}
ggplot(d) + geom_bar(aes(year, all.cs), stat = "identity", fill = "blue") + geom_bar(aes(year, free.cs), stat = "identity", fill = "orange") + theme_bw() + ylab("cumulative sum of articles") + xlab("year") + scale_x_continuous(breaks = seq(2000, 2013, by = 2)) + labs(title = "Cumulative sum of PubMed articles 2000-2013\nblue = all; orange = freetext")
```

## Cumulative increase in articles relative to start year

The rate of accumulation of free full text articles compared with the base year of 2000 is somewhat higher than that for all articles.

```{r tidy=TRUE, cache=TRUE}
d2 <- data.frame(year = 2000:2013, all = d$all.cs / d$all.cs[1], free = d$free.cs / d$free.cs[1])
d2.m <- melt(d2, id.vars = "year")
ggplot(d2.m) + geom_point(aes(year, value, color = variable)) + theme_bw() + scale_color_manual(values = c("blue", "orange")) + scale_x_continuous(breaks = seq(2000, 2013, by = 2)) + ylab("cumulative sum of articles relative to 2000") + labs(title = "Cumulative sum of PubMed articles 2000-2013 relative to 2000\nblue = all; orange = freetext")
```

## Change in number of articles as a proportion of previous year

The year-on-year change in number of articles is not very informative, as there is a lot of unexplained fluctuation.

```{r cache=TRUE, tidy=TRUE}
d3 <- data.frame(year = NA, all = NA, free = NA)
for(i in 1:13) {
    d3[i, "year"] <- d[i+1, "year"]
    d3[i, "all"]  <- (d[i+1, "all"] - d[i, "all"]) / d[i, "all"]
    d3[i, "free"] <- (d[i+1, "free"] - d[i, "free"]) / d[i, "free"]
}
d3.m <- melt(d3, id.vars = "year")
ggplot(d3.m) + geom_line(aes(year, value, color = variable)) + theme_bw() + scale_color_manual(values = c("blue", "orange")) + scale_x_continuous(breaks = seq(2000, 2013, by = 2)) + ylab("relative change in number of articles") + labs(title = "Change in number of PubMed articles relative to previous year\nblue = all; orange = freetext")
```

# Conclusions

From 2000-2013:

* Total and freetext articles in PubMed rose steadily
* The proportion of freetext articles rose from around 21% to 33%
* Total articles accumulated at a higher rate than freetext articles
* From 2005 onwards, freetext articles accumulated at a slightly higher rate than total articles relative to a baseline year of 2000
* Year-on-year changes in article numbers show no clear pattern
* ~~Number of articles in PMC is not a good proxy for number of open-access articles~~

