---
title: "pmRetract PubMed retractions report"
author: "Neil Saunders"
date: "compiled `r Sys.time()`"
output: html_document
---

```{r echo=FALSE, ref.label="setup", message=FALSE}
```

## Introduction
This is the latest report on retracted publications in the [PubMed database](http://www.pubmed.org). It was generated on `r 
strftime(Sys.Date(), "%B %d %Y")`. This report replaces the application PMRetract, formerly hosted at [Heroku](https://www.heroku.com).

## 1. Timeline of retractions
This chart shows the number of retracted publications per year. PubMed uses a variety of different dates; the year used here is the date that the record was created (CRDT).

Clicking on a year will open a new window at PubMed showing the retracted articles for that year.

```{r echo=FALSE, ref.label="timeline", message=FALSE}
```

```{r chart1, message=FALSE, echo=FALSE, results='asis', warning=FALSE}
hc1 <- plotTimeline(doc)
hc1$show('inline', cdn = FALSE, include_assets = TRUE)
```

## 2. Cumulative timeline of retractions
This chart shows the cumulative sum of retracted publications per year. The year used here is the date that the record was created (CRDT).

Clicking on a year will open a new window at PubMed showing the retracted articles from 1959 up to and including that year.

```{r echo=FALSE, ref.label="cumulative", message=FALSE}
```

```{r chart2, message=FALSE, echo=FALSE, results='asis', warning=FALSE}
hc2 <- plotCumSumTimeline(doc)
hc2$show('inline', cdn = FALSE, include_assets = TRUE)
```

## 3. Retraction rate by year
This chart shows the rate of retracted publications per year, as retractions per 100 000 publications. The year used here is the date that the record was created (CRDT).

Clicking on a year will open a new window at PubMed showing the retracted articles for that year.

```{r echo=FALSE, ref.label="byyear", message=FALSE}
```

```{r chart3, message=FALSE, echo=FALSE, results='asis', warning=FALSE}
hc3 <- plotByYear(doc, years.total)
hc3$show('inline', cdn = FALSE, include_assets = TRUE)
```

## 4. Retractions by Journal
This chart shows the top 20 journals by number of retracted articles.

Clicking on a journal name will open a new window at PubMed showing the retracted articles from that journal.

```{r echo=FALSE, ref.label="byjournal", message=FALSE}
```

```{r chart4, message=FALSE, echo=FALSE, results='asis', warning=FALSE}
hc4 <- plotByJournal(doc)
hc4$show('inline', cdn = FALSE, include_assets = TRUE)
```

## 5. Functions

### 5.1 Setup
This code loads required libraries and pre-saved data.

```{r setup, tidy=TRUE, warning=FALSE, eval=FALSE}
library(rCharts)
library(rentrez)
library(XML)

setwd("../../data")
doc <- xmlTreeParse("retracted.xml", useInternalNodes = TRUE)
years.total <- read.csv("years.csv")
```

### 5.2 Timeline
This code generates the timeline chart.

```{r timeline, tidy=TRUE, warning=FALSE, eval=FALSE}
plotTimeline <- function(d) {
  dates <- xpathSApply(d, "//PubmedData/History/PubMedPubDate[@PubStatus='entrez']/Year", xmlValue)
  df1 <- as.data.frame(table(dates), stringsAsFactors = FALSE)
  hc <- Highcharts$new()
  hc$title(text = "Retracted publications by year of Entrez record creation")
  hc$series(data = df1$Freq, type = "column")
  hc$xAxis(categories = df1$dates, labels = list(rotation = 90, formatter = "#! function() { return '<a href=\"http://www.pubmed.org/?term=%22Retracted Publication%22[PTYP] AND ' + escape(this.value) + '[CRDT]\" target=\"_blank\">' + this.value + '</a>'; } !#", useHTML = "true"), title = list(text = "year"))
  hc$yAxis(title = list(text = "retracted publications"))
  hc$legend(enabled = FALSE)
  hc$tooltip(pointFormat = "{point.y} records")
  return(hc)
}
```

### 5.3 Cumulative timeline
This code generates the cumulative timeline chart.

```{r cumulative, tidy=TRUE, warning=FALSE, eval=FALSE}
plotCumSumTimeline <- function(d) {
  dates <- xpathSApply(d, "//PubmedData/History/PubMedPubDate[@PubStatus='entrez']/Year", xmlValue)
  df1 <- as.data.frame(table(dates), stringsAsFactors = FALSE)
  hc <- Highcharts$new()
  hc$title(text = "Cumulative sum of retracted publications by year of Entrez record creation")
  hc$series(data = cumsum(df1$Freq), type = "column")
  hc$xAxis(categories = df1$dates, labels = list(rotation = 90, formatter = "#! function() { return '<a href=\"http://www.pubmed.org/?term=%22Retracted Publication%22[PTYP] AND 1959:' + escape(this.value) + '[CRDT]\" target=\"_blank\">' + this.value + '</a>'; } !#", useHTML = "true"), title = list(text = "year"))
  hc$yAxis(title = list(text = "sum of retracted publications"))
  hc$legend(enabled = FALSE)
  hc$tooltip(pointFormat = "{point.y} records since 1959")
  return(hc)
}
```

### 5.4 By year
This code generates the retraction rate by year chart.

```{r byyear, tidy=TRUE, warning=FALSE, eval=FALSE}
plotByYear <- function(d, total) {
  dates <- xpathSApply(d, "//PubmedData/History/PubMedPubDate[@PubStatus='entrez']/Year", xmlValue)
  years <- as.numeric(dates)
  ydf <- data.frame(year = min(years):max(years), total = NA, retracted = NA)
  years.cnt <- as.data.frame(table(years), stringsAsFactors = FALSE)
  m <- match(ydf$year, years.cnt$years)
  ydf$retracted <- years.cnt[m, "Freq"]
  ydf$retracted <- ifelse(is.na(ydf$retracted), 0, ydf$retracted)
  m <- match(ydf$year, total$year)
  ydf$total <- total[m, "total"]  
  hc <- Highcharts$new()
  hc$title(text = "Retracted publications per 100 000 publications by year of Entrez record creation")
  hc$series(data = as.numeric(sprintf("%.3f", (100000 / ydf$total) * ydf$retracted)), type = "column", events = list(click = "#! function() {window.open(this.options.url)} !#"))
  hc$xAxis(categories = ydf$year, labels = list(rotation = 90, formatter = "#! function() { return '<a href=\"http://www.pubmed.org/?term=%22Retracted Publication%22[PTYP] AND ' + escape(this.value) + '[CRDT]\" target=\"_blank\">' + this.value + '</a>'; } !#", useHTML = "true"), title = list(text = "year"))
  hc$yAxis(title = list(text = "retracted publications per 100 000 publications"))
  hc$legend(enabled = FALSE)
  hc$tooltip(pointFormat = "{point.y} retracted records per 100 000 publication records")
  return(hc)
}
```

### 5.5 By journal
This code generates the retractions by journal chart.

```{r byjournal, tidy=TRUE, warning=FALSE, eval=FALSE}
plotByJournal <- function(d) {
  journals <- xpathSApply(d, "//MedlineCitation/Article/Journal/ISOAbbreviation", xmlValue)
  journals.cnt <- as.data.frame(table(journals), stringsAsFactors = FALSE)
  colnames(journals.cnt) <- c("journal", "count")
  j20 <- head(journals.cnt[order(journals.cnt$count, decreasing = TRUE),], 20)
  hc <- Highcharts$new()
  hc$chart(marginLeft = 220)
  hc$series(data = j20$count, type = "bar")
  hc$xAxis(categories = j20$journal, labels = list(formatter = "#! function() { return '<a href=\"http://www.pubmed.org/?term=%22Retracted Publication%22[PTYP] AND %22' + escape(this.value) + '%22[JOUR]\" target=\"_blank\">' + this.value + '</a>'; } !#", useHTML = "true"))
  hc$yAxis(title = list(text = "retracted publications"))
  hc$legend(enabled = FALSE)
  hc$tooltip(pointFormat = "{point.y} records")
  return(hc)
}
