---
title: "PubMed retractions report"
author: "Neil Saunders"
date: "compiled `r Sys.time()`"
output:
  html_document:
    highlight: pygments
    keep_md: yes
    theme: cerulean
    number_sections: yes
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

library(tidyverse)
library(highcharter)
library(xml2)

doc.retd    <- read_xml("../../data/retracted.xml")
doc.retOf   <- read_xml("../../data/retractionOf.xml")
years.total <- read_csv("../../data/years.csv")
jour.retd   <- read_csv("../../data/journals_retracted.csv")
jour.retOf  <- read_csv("../../data/journals_retractionOf.csv")
```

# Introduction
This is the latest report on retracted publications in the [PubMed database](http://www.pubmed.gov). It was generated on `r 
strftime(Sys.Date(), "%B %d %Y")`.

Each section contains two charts. The first (blue) contains data about _retracted publications_. The second (orange) contains data about _retraction notices_. The PubMed search terms for these are, respectively:

* "retracted publication"[PTYP]
* "retraction of publication"[PTYP]

# Timeline of retractions
This chart shows the number of retracted publications per year. PubMed uses a variety of different dates; the year used here is the date that the record was created (CRDT).

Clicking on a year will open a new window at PubMed showing the retracted articles for that year.

## Retracted Publications
```{r chart1-1}
# create data frame
dat <- doc.retd %>% 
  xml_find_all("//PubmedData/History/PubMedPubDate[@PubStatus='entrez']/Year") %>% 
  xml_text() %>% 
  as_tibble() %>% 
  mutate(year = as.numeric(value)) %>% 
  count(year) %>% 
  right_join(tibble(year = min(.$year):max(.$year))) %>% 
  arrange(year)

# plot chart
highchart() %>% 
  hc_add_series(dat, "column", hcaes(x = year, y = n)) %>%
  hc_legend(enabled = FALSE) %>%
  hc_title(text = "Retracted publications by year of Entrez record creation") %>%
  hc_xAxis(title = list(text = "year"),
           labels = list(rotation = 90,
                         formatter = JS("function() { return '<a href=\"http://www.pubmed.gov/?term=%22retracted%20publication%22[PTYP] AND ' + escape(this.value) + '[CRDT]\" target=\"_blank\">' + this.value + '</a>'; }"),
                         useHTML = TRUE,
                         style = list(fontSize = "10px")),
           tickInterval = 1,
           type = "datetime") %>%
  hc_yAxis(title = list(text = "retracted publications")) %>%
  hc_tooltip(pointFormat = '{point.y} records')
```

## Retraction Notices
```{r chart1-2}
# create data frame
dat <- doc.retOf %>% 
  xml_find_all("//PubmedData/History/PubMedPubDate[@PubStatus='entrez']/Year") %>% 
  xml_text() %>% 
  as_tibble() %>% 
  mutate(year = as.numeric(value)) %>% 
  count(year) %>% 
  right_join(tibble(year = min(.$year):max(.$year))) %>% 
  arrange(year)

# plot chart
highchart() %>% 
  hc_add_series(dat, "column", hcaes(x = year, y = n)) %>%
  hc_legend(enabled = FALSE) %>%
  hc_title(text = "Retraction notices by year of Entrez record creation") %>%
  hc_xAxis(title = list(text = "year"),
           labels = list(rotation = 90,
                         formatter = JS("function() { return '<a href=\"http://www.pubmed.gov/?term=%22retraction%20of%20publication%22[PTYP] AND ' + escape(this.value) + '[CRDT]\" target=\"_blank\">' + this.value + '</a>'; }"),
                         useHTML = TRUE,
                         style = list(fontSize = "10px")),
           tickInterval = 1,
           type = "datetime") %>%
  hc_yAxis(title = list(text = "retraction notices")) %>%
  hc_tooltip(pointFormat = '{point.y} records') %>%
  hc_plotOptions(series = list(color = "#FFA500"))
```

# Cumulative timeline of retractions
This chart shows the cumulative sum of retracted publications per year. The year used here is the date that the record was created (CRDT).

Clicking on a year will open a new window at PubMed showing the retracted articles from 1959 up to and including that year.

## Retracted Publications
```{r chart2-1}
# create data frame
dat <- doc.retd %>% 
  xml_find_all("//PubmedData/History/PubMedPubDate[@PubStatus='entrez']/Year") %>% 
  xml_text() %>% 
  as_tibble() %>% 
  mutate(year = as.numeric(value)) %>% 
  count(year) %>% 
  right_join(tibble(year = min(.$year):max(.$year))) %>% 
  arrange(year) %>% 
  replace_na(list(n = 0)) %>%
  mutate(cs = cumsum(n))

# plot chart
highchart() %>% 
  hc_add_series(dat, "column", hcaes(x = year, y = cs)) %>%
  hc_legend(enabled = FALSE) %>%
  hc_title(text = "Cumulative sum of retracted publications by year of Entrez record creation") %>%
  hc_xAxis(title = list(text = "year"),
           labels = list(rotation = 90,
                         formatter = JS("function() { return '<a href=\"http://www.pubmed.gov/?term=%22retracted%20publication%22[PTYP] AND 1959:' + escape(this.value) + '[CRDT]\" target=\"_blank\">' + this.value + '</a>'; }"),
                         useHTML = TRUE,
                         style = list(fontSize = "10px")),
           tickInterval = 1,
           type = "datetime") %>%
  hc_yAxis(title = list(text = "sum of retracted publications")) %>%
  hc_tooltip(pointFormat = '{point.y} records since 1959')
```

## Retraction Notices
```{r chart2-2}
# create data frame
dat <- doc.retOf %>% 
  xml_find_all("//PubmedData/History/PubMedPubDate[@PubStatus='entrez']/Year") %>% 
  xml_text() %>% 
  as_tibble() %>% 
  mutate(year = as.numeric(value)) %>% 
  count(year) %>% 
  right_join(tibble(year = min(.$year):max(.$year))) %>% 
  arrange(year) %>% 
  replace_na(list(n = 0)) %>%
  mutate(cs = cumsum(n))

# plot chart
highchart() %>% 
  hc_add_series(dat, "column", hcaes(x = year, y = cs)) %>%
  hc_legend(enabled = FALSE) %>%
  hc_title(text = "Cumulative sum of retraction notices by year of Entrez record creation") %>%
  hc_xAxis(title = list(text = "year"),
           labels = list(rotation = 90,
                         formatter = JS("function() { return '<a href=\"http://www.pubmed.gov/?term=%22retraction%20of%20publication%22[PTYP] AND 1959:' + escape(this.value) + '[CRDT]\" target=\"_blank\">' + this.value + '</a>'; }"),
                         useHTML = TRUE,
                         style = list(fontSize = "10px")),
           tickInterval = 1,
           type = "datetime") %>%
  hc_yAxis(title = list(text = "sum of retraction notices")) %>%
  hc_tooltip(pointFormat = '{point.y} records since 1959') %>%
  hc_plotOptions(series = list(color = "#FFA500"))
```

# Retraction rate by year
This chart shows the rate of retracted publications per year, as retractions per 100 000 publications. The year used here is the date that the record was created (CRDT).

Clicking on a year will open a new window at PubMed showing the retracted articles for that year.

## Retracted Publications
```{r chart3-1}
# create data frame
dat <- doc.retd %>% 
  xml_find_all("//PubmedData/History/PubMedPubDate[@PubStatus='entrez']/Year") %>% 
  xml_text() %>% 
  as_tibble() %>% 
  mutate(year = as.numeric(value)) %>% 
  count(year) %>% 
  right_join(tibble(year = min(.$year):max(.$year))) %>% 
  right_join(years.total) %>% 
  mutate(rate = round((100000/total) * n, 3)) %>% 
  arrange(year)

# plot chart
highchart() %>% 
  hc_add_series(dat, "column", hcaes(x = year, y = rate)) %>%
  hc_legend(enabled = FALSE) %>%
  hc_title(text = "Retracted publications per 100 000 publications by year of Entrez record creation") %>%
  hc_xAxis(labels = list(rotation = 90,
                         formatter = JS("function() { return '<a href=\"http://www.pubmed.gov/?term=%22retracted%20publication%22[PTYP] AND ' + escape(this.value) + '[CRDT]\" target=\"_blank\">' + this.value + '</a>'; }"),
                         useHTML = "true",
                         style = list(fontSize = "10px")),
           title = list(text = "year"),
           tickInterval = 1,
           type = "datetime") %>%
  hc_yAxis(title = list(text = "retracted publications/100 000 publications")) %>%
  hc_tooltip(pointFormat = "{point.y} retracted records per 100 000 publication records")
```

## Retraction Notices
```{r chart3-2}
# create data frame
dat <- doc.retOf %>% 
  xml_find_all("//PubmedData/History/PubMedPubDate[@PubStatus='entrez']/Year") %>% 
  xml_text() %>% 
  as_tibble() %>% 
  mutate(year = as.numeric(value)) %>% 
  count(year) %>% 
  right_join(tibble(year = min(.$year):max(.$year))) %>% 
  right_join(years.total) %>% 
  mutate(rate = round((100000/total) * n, 3)) %>% 
  arrange(year)

# plot chart
highchart() %>% 
  hc_add_series(dat, "column", hcaes(x = year, y = rate)) %>%
  hc_legend(enabled = FALSE) %>%
  hc_title(text = "Retraction notices per 100 000 publications by year of Entrez record creation") %>%
  hc_xAxis(labels = list(rotation = 90,
                         formatter = JS("function() { return '<a href=\"http://www.pubmed.gov/?term=%22retraction%20of%20publication%22[PTYP] AND ' + escape(this.value) + '[CRDT]\" target=\"_blank\">' + this.value + '</a>'; }"),
                         useHTML = "true",
                         style = list(fontSize = "10px")), 
           title = list(text = "year"),
           tickInterval = 1,
           type = "datetime") %>%
  hc_yAxis(title = list(text = "retraction notices/100 000 publications")) %>%
  hc_tooltip(pointFormat = "{point.y} retraction notices per 100 000 publication records") %>%
  hc_plotOptions(series = list(color = "#FFA500"))
```

# Retractions by journal (absolute numbers)
This chart shows the top 20 journals by number of retracted articles. See section 5 for journals ranked by retractions relative to total articles.

Clicking on a journal name will open a new window at PubMed showing the retracted articles from that journal.

## Retracted Publications
```{r chart4-1}
# create data frame
dat <- jour.retd %>%
  top_n(20, wt = count) %>% 
  arrange(desc(count))

# plot chart
highchart() %>% 
  hc_add_series(dat, "bar", hcaes(x = journal, y = count)) %>%
  hc_legend(enabled = FALSE) %>%
  hc_yAxis(title = list(text = "retractions")) %>%
  hc_xAxis(categories = dat$journal,
           labels = list(formatter = JS("function() { return '<a href=\"http://www.pubmed.gov/?term=%22retracted%20publication%22[PTYP] AND %22' + escape(this.value) + '%22[JOUR]\" target=\"_blank\">' + this.value + '</a>'; }"),
                         useHTML = "true",
                         style = list(fontSize = "10px"))) %>%
  hc_tooltip(pointFormat = "{point.y} records")
```

## Retraction Notices
```{r chart4-2}
# create data frame
dat <- jour.retOf %>%
  top_n(20, wt = count) %>% 
  arrange(desc(count))

# plot chart
highchart() %>% 
  hc_add_series(dat, "bar", hcaes(x = journal, y = count)) %>%
  hc_legend(enabled = FALSE) %>%
  hc_yAxis(title = list(text = "retraction notices")) %>%
  hc_xAxis(categories = dat$journal,
           labels = list(formatter = JS("function() { return '<a href=\"http://www.pubmed.gov/?term=%22retraction%20of%20publication%22[PTYP] AND %22' + escape(this.value) + '%22[JOUR]\" target=\"_blank\">' + this.value + '</a>'; }"),
                         useHTML = "true",
                         style = list(fontSize = "10px"))) %>%
  hc_tooltip(pointFormat = "{point.y} records") %>%
  hc_plotOptions(series = list(color = "#FFA500"))
```

# Retractions by journal (normalised)
This chart ranks the top 20 retractions by journal. For each journal, retractions (or retraction notices) per 100 000 publications from that journal are shown.

Clicking on a journal name will open a new window at PubMed showing the retracted articles from that journal.

## Retracted Publications
```{r chart5-1}
# create data frame
dat <- jour.retd %>%
  mutate(rate = round((100000/total) * count, 3)) %>%
  top_n(20, wt = rate) %>% 
  arrange(desc(rate))

# plot chart
highchart() %>% 
  hc_add_series(dat, "bar", hcaes(x = journal, y = rate)) %>%
  hc_legend(enabled = FALSE) %>%
  hc_yAxis(title = list(text = "retractions")) %>%
  hc_xAxis(categories = dat$journal,
           labels = list(formatter = JS("function() { return '<a href=\"http://www.pubmed.gov/?term=%22retracted%20publication%22[PTYP] AND %22' + escape(this.value) + '%22[JOUR]\" target=\"_blank\">' + this.value + '</a>'; }"), 
                         useHTML = "true",
                         style = list(fontSize = "10px"))) %>%
  hc_tooltip(pointFormat = "{point.y} retractions / 100 000 publications")
```

## Retraction Notices
```{r chart5-2}
# create data frame
dat <- jour.retOf %>%
  mutate(rate = round((100000/total) * count, 3)) %>%
  top_n(20, wt = rate) %>% 
  arrange(desc(rate))

# plot chart
highchart() %>% 
  hc_add_series(dat, "bar", hcaes(x = journal, y = rate)) %>%
  hc_legend(enabled = FALSE) %>%
  hc_yAxis(title = list(text = "retraction notices")) %>%
  hc_xAxis(categories = dat$journal,
           labels = list(formatter = JS("function() { return '<a href=\"http://www.pubmed.gov/?term=%22retraction%20of%20publication%22[PTYP] AND %22' + escape(this.value) + '%22[JOUR]\" target=\"_blank\">' + this.value + '</a>'; }"), 
                         useHTML = "true",
                         style = list(fontSize = "10px"))) %>%
  hc_tooltip(pointFormat = "{point.y} retraction notices / 100 000 publications") %>%
  hc_plotOptions(series = list(color = "#FFA500"))
```
