---
title: "PubMed retractions report"
author: "Neil Saunders"
date: "compiled `r Sys.time()`"
output: html_document
---

```{r echo=FALSE, ref.label="setup", message=FALSE}
```

## Introduction
This is the latest report on retracted publications in the [PubMed database](http://www.pubmed.org). It was generated on `r 
strftime(Sys.Date(), "%B %d %Y")`. This report replaces the application PMRetract, formerly hosted at [Heroku](https://www.heroku.com).

Each section contains two charts. The first (blue) contains data about _retracted publications_. The second (orange) contains data about _retraction notices_. The PubMed search terms for these are, respectively:

* "retracted publication"[PTYP]
* "retraction of publication"[PTYP]

## 1. Timeline of retractions
This chart shows the number of retracted publications per year. PubMed uses a variety of different dates; the year used here is the date that the record was created (CRDT).

Clicking on a year will open a new window at PubMed showing the retracted articles for that year.

```{r echo=FALSE, ref.label="timeline", message=FALSE}
```

### 1.1 Retracted Publications
```{r chart1-1, message=FALSE, echo=FALSE, results='asis', warning=FALSE}
hc1.1 <- plotTimeline(doc.retd, "retracted publication")
hc1.1$show('inline', cdn = FALSE, include_assets = TRUE)
```

### 1.2 Retraction Notices
```{r chart1-2, message=FALSE, echo=FALSE, results='asis', warning=FALSE}
hc1.2 <- plotTimeline(doc.retOf, "retraction of publication")
hc1.2$title(text = "Retraction notices by year of Entrez record creation")
hc1.2$yAxis(title = list(text = "retraction notices"))
hc1.2$plotOptions(series = list(color = "#FFA500"))
hc1.2$show('inline', cdn = FALSE, include_assets = TRUE)
```

## 2. Cumulative timeline of retractions
This chart shows the cumulative sum of retracted publications per year. The year used here is the date that the record was created (CRDT).

Clicking on a year will open a new window at PubMed showing the retracted articles from 1959 up to and including that year.

```{r echo=FALSE, ref.label="cumulative", message=FALSE}
```

### 2.1 Retracted Publications
```{r chart2-1, message=FALSE, echo=FALSE, results='asis', warning=FALSE}
hc2.1 <- plotCumSumTimeline(doc.retd, "retracted publication")
hc2.1$show('inline', cdn = FALSE, include_assets = TRUE)
```

### 2.2 Retraction Notices
```{r chart2-2, message=FALSE, echo=FALSE, results='asis', warning=FALSE}
hc2.2 <- plotCumSumTimeline(doc.retOf, "retraction of publication")
hc2.2$title(text = "Cumulative sum of retraction notices by year of Entrez record creation")
hc2.2$yAxis(title = list(text = "sum of retraction notices"))
hc2.2$plotOptions(series = list(color = "#FFA500"))
hc2.2$show('inline', cdn = FALSE, include_assets = TRUE)
```

## 3. Retraction rate by year
This chart shows the rate of retracted publications per year, as retractions per 100 000 publications. The year used here is the date that the record was created (CRDT).

Clicking on a year will open a new window at PubMed showing the retracted articles for that year.

```{r echo=FALSE, ref.label="byyear", message=FALSE}
```

### 3.1 Retracted Publications
```{r chart3-1, message=FALSE, echo=FALSE, results='asis', warning=FALSE}
hc3.1 <- plotByYear(doc.retd, years.total, "retracted publication")
hc3.1$show('inline', cdn = FALSE, include_assets = TRUE)
```

### 3.2 Retraction Notices
```{r chart3-2, message=FALSE, echo=FALSE, results='asis', warning=FALSE}
hc3.2 <- plotByYear(doc.retOf, years.total, "retraction of publication")
hc3.2$title(text = "Retraction notices per 100 000 publications by year of Entrez record creation")
hc3.2$yAxis(title = list(text = "retraction notices per 100 000 publications"))
hc3.2$tooltip(pointFormat = "{point.y} retraction notices per 100 000 publication records")
hc3.2$plotOptions(series = list(color = "#FFA500"))
hc3.2$show('inline', cdn = FALSE, include_assets = TRUE)
```

## 4. Retractions by journal
This chart shows the top 20 journals by number of retracted articles. Ideally, these should be normalised to take into account the total number of articles per journal. However, this hardly seems worthwhile given the relatively-low number of retractions (tens) versus total publications (hundreds of thousands). Keep those issues in mind when interpreting these charts.

Clicking on a journal name will open a new window at PubMed showing the retracted articles from that journal.

```{r echo=FALSE, ref.label="byjournal", message=FALSE}
```

### 4.1 Retracted Publications
```{r chart4-1, message=FALSE, echo=FALSE, results='asis', warning=FALSE}
hc4.1 <- plotByJournal(doc.retd, "retracted publication")
hc4.1$show('inline', cdn = FALSE, include_assets = TRUE)
```

### 4.2 Retraction Notices
```{r chart4-2, message=FALSE, echo=FALSE, results='asis', warning=FALSE}
hc4.2 <- plotByJournal(doc.retOf, "retraction of publication")
hc4.2$yAxis(title = list(text = "retraction notices"))
hc4.2$plotOptions(series = list(color = "#FFA500"))
hc4.2$show('inline', cdn = FALSE, include_assets = TRUE)
```

## 5. Functions

### 5.1 Setup
This code loads required libraries and pre-saved data.

```{r setup, tidy=TRUE, warning=FALSE, eval=FALSE}
library(rCharts)
library(rentrez)
library(XML)

setwd("../../data")
doc.retd <- xmlTreeParse("retracted.xml", useInternalNodes = TRUE)
doc.retOf <- xmlTreeParse("retractionOf.xml", useInternalNodes = TRUE)
years.total <- read.csv("years.csv")
```

### 5.2 Timeline
This code generates the timeline chart.

```{r timeline, tidy=TRUE, warning=FALSE, eval=FALSE}
plotTimeline <- function(d, term) {
  dates <- xpathSApply(d, "//PubmedData/History/PubMedPubDate[@PubStatus='entrez']/Year", xmlValue)
  df1 <- as.data.frame(table(dates), stringsAsFactors = FALSE)
  hc <- Highcharts$new()
  hc$title(text = "Retracted publications by year of Entrez record creation")
  hc$series(data = df1$Freq, type = "column")
  hc$xAxis(categories = df1$dates, labels = list(rotation = 90, formatter = paste("#! function() { return '<a href=\"http://www.pubmed.org/?term=%22", term, "%22[PTYP] AND ' + escape(this.value) + '[CRDT]\" target=\"_blank\">' + this.value + '</a>'; } !#", sep = ""), useHTML = "true"), title = list(text = "year"))
  hc$yAxis(title = list(text = "retracted publications"))
  hc$legend(enabled = FALSE)
  hc$tooltip(pointFormat = "{point.y} records")
  return(hc)
}
```

### 5.3 Cumulative timeline
This code generates the cumulative timeline chart.

```{r cumulative, tidy=TRUE, warning=FALSE, eval=FALSE}
plotCumSumTimeline <- function(d, term) {
  dates <- xpathSApply(d, "//PubmedData/History/PubMedPubDate[@PubStatus='entrez']/Year", xmlValue)
  df1 <- as.data.frame(table(dates), stringsAsFactors = FALSE)
  hc <- Highcharts$new()
  hc$title(text = "Cumulative sum of retracted publications by year of Entrez record creation")
  hc$series(data = cumsum(df1$Freq), type = "column")
  hc$xAxis(categories = df1$dates, labels = list(rotation = 90, formatter = paste("#! function() { return '<a href=\"http://www.pubmed.org/?term=%22", term, "%22[PTYP] AND 1959:' + escape(this.value) + '[CRDT]\" target=\"_blank\">' + this.value + '</a>'; } !#", sep = ""), useHTML = "true"), title = list(text = "year"))
  hc$yAxis(title = list(text = "sum of retracted publications"))
  hc$legend(enabled = FALSE)
  hc$tooltip(pointFormat = "{point.y} records since 1959")
  return(hc)
}
```

### 5.4 By year
This code generates the retraction rate by year chart.

```{r byyear, tidy=TRUE, warning=FALSE, eval=FALSE}
plotByYear <- function(d, total, term) {
  dates <- xpathSApply(d, "//PubmedData/History/PubMedPubDate[@PubStatus='entrez']/Year", xmlValue)
  years <- as.numeric(dates)
  ydf <- data.frame(year = min(years):max(years), total = NA, retracted = NA)
  years.cnt <- as.data.frame(table(years), stringsAsFactors = FALSE)
  m <- match(ydf$year, years.cnt$years)
  ydf$retracted <- years.cnt[m, "Freq"]
  ydf$retracted <- ifelse(is.na(ydf$retracted), 0, ydf$retracted)
  m <- match(ydf$year, total$year)
  ydf$total <- total[m, "total"]  
  hc <- Highcharts$new()
  hc$title(text = "Retracted publications per 100 000 publications by year of Entrez record creation")
  hc$series(data = as.numeric(sprintf("%.3f", (100000 / ydf$total) * ydf$retracted)), type = "column", events = list(click = "#! function() {window.open(this.options.url)} !#"))
  hc$xAxis(categories = ydf$year, labels = list(rotation = 90, formatter = paste("#! function() { return '<a href=\"http://www.pubmed.org/?term=%22", term, "%22[PTYP] AND ' + escape(this.value) + '[CRDT]\" target=\"_blank\">' + this.value + '</a>'; } !#", sep = ""), useHTML = "true"), title = list(text = "year"))
  hc$yAxis(title = list(text = "retracted publications per 100 000 publications"))
  hc$legend(enabled = FALSE)
  hc$tooltip(pointFormat = "{point.y} retracted records per 100 000 publication records")
  return(hc)
}
```

### 5.5 By journal
This code generates the retractions by journal chart.

```{r byjournal, tidy=TRUE, warning=FALSE, eval=FALSE}
plotByJournal <- function(d, term) {
  journals <- xpathSApply(d, "//MedlineCitation/Article/Journal/ISOAbbreviation", xmlValue)
  journals.cnt <- as.data.frame(table(journals), stringsAsFactors = FALSE)
  colnames(journals.cnt) <- c("journal", "count")
  j20 <- head(journals.cnt[order(journals.cnt$count, decreasing = TRUE),], 20)
  hc <- Highcharts$new()
  hc$chart(marginLeft = 220)
  hc$series(data = j20$count, type = "bar")
  hc$xAxis(categories = j20$journal, labels = list(formatter = paste("#! function() { return '<a href=\"http://www.pubmed.org/?term=%22", term, "%22[PTYP] AND %22' + escape(this.value) + '%22[JOUR]\" target=\"_blank\">' + this.value + '</a>'; } !#", sep = ""), useHTML = "true"))
  hc$yAxis(title = list(text = "retracted publications"))
  hc$legend(enabled = FALSE)
  hc$tooltip(pointFormat = "{point.y} records")
  return(hc)
}
